Performance benefits of Intel Optane DC storage for the parallel processing of large neuroimaging data

### 2019-09-19 report/report.tex
addressed @glatard's comments

\begin{abstract}
    Open-access neuroimaging datasets have reached up to $\SI{}{\peta\byte}$ of data, and
    continue to grow. The ability to leverage the entirety of these datasets are
    limited to a restricted number of labs with both the capacity and infrastructure
    to process the data. Whereas BigData engines have significantly reduced
    application performance penalties with respect to data movement, their applied
    strategies are not necessarily practical within neuroimaging workflows where
    intermediary results may need to be materialized to shared storage for post-processing
    analysis. In this paper we evaluate the performance advantage brought by Intel
    Optane DC \tristan{expan acronym?} persistent memory storage for the processing of large neuroimaging 
    datasets using the two available configurations modes: Memory Mode and App Direct Mode.
    We employ a synthetic algorithm on the $\SI{76}{\gibi\byte}$ \bigbrain
    and a real neuroimaging application on the CoRR \tristan{expand acronym} dataset using 25 and 96 parallel
    processes in both cases. Our results show that performance of all available storage
    devices (e.g. tmpfs, local disk and Isilon) is somewhat deteriorated by memory mode
    when compared to App Direct Mode. In contrast, Optane in App Direct Mode was
    found to be slower than local disk, when the load on the bus was high due to sheer
    amount of concurrent processes and short task durations. It is believed that
    Optane would likely be useful in the accessing of typically larger-than-memory
    data in a limited-parallelism manner, such as with visualization applications.
    Further experiments with larger datasets remain to be performed.

@@

Due to storage limitations, only subsets of such neuroimaging repositories 
can be processed in a typical research laboratory. Moreover, as these datasets 
are extremely large and are only increasing in size, they are typically stored 
in higher-capacity, slower storage devices \tristan{such as?}. In such conditions, large-scale 
studies in neuroimaging remain limited to labs with access to adequate 
infrastructure and are limited by data transfer times.

Furthermore, intermediary data is often required during results analysis. This limits
any performance benefits that can arise from volatile in-memory computing as intermediary
results need to be materialized onto persistent storage. As neuroimaging datasets continue
to increase in size, this movement of data will significantly increase the processing time
of neuroimaging data.

Whereas popular Big Data engines, such as MapReduce~\cite{mapreduce} and Apache Spark~\cite{spark}, have
incorporated software solutions to limit data transfers (e.g. in-memory computing,
data locality, and lazy evaluation), hardware has also adapted to the growing datasets.
One such improvement is the concept of placing flash storage directly on the Dual
In-line Memory Module (DIMM), thereby reducing the latency of accessing data on 
flash storage devices. While the latency of these devices is improved, the bandwidth
remains the same. However, as noted by \cite{nvdimms}, a severe performance degradation
can be experienced by having memory traffic and I/O placed on the same bus \tristan{references are
missing or the bib file hasn't been updated}.

Intel Optane DC Persistent Memory storagei~\cite{optanebrief} is a recently introduced high-performance
storage technology that also sits on the DIMMs to reduce latency to the device.
Unlike DRAM DIMMs, it can accommodate twice the volume ($\SI{512}{\giga\byte}$) of
storage enabling it to be a more cost effective alternative to DRAM storage.
There are two principle configuration modes, namely Memory Mode and App Direct Mode,
that enable the storage to either be accessed as an extension of available memory 
or as a non-volatile memory storage device.

@@

\tristan{awesome intro Valerie, great context, right level of detail, nice 
parallel between software and hardware solutions for big data, super well informed.}

@@

The server used in our experiments contain $\SI{3}{\tera\byte}$ of Intel Optane Persistent memory, 768GB of DRAM,
Isilon, local disk(SSD).
CPU: Cascade lake
OS: Red Hat 7.1
\valerie{include storage bandwidth table here}

@@

\subsubsection{App Direct Mode}

App Direct Mode enables Optane to be accessed as a high-performance storage device.
Unlike Memory Mode, the OS is able to differentiate between main memory and Optane
storage accesses, treating them as two distinct entities. As Optane is located on
the DIMMs, it exhibits lower latencies than traditional SSDs, while being able
to be accessed like an SSD.

\tristan{I guess tmpfs, Isilon and local disks should be described here.}

@@

Using this inequality, we anticipate that whenever the application is within the limits,
the storage selection will make little difference as storage devices with writeback enabled \tristan{This should be explained 
before, in methods (infrastructure) and possibly in intro.}
will be able to leverage the memory cache. It is expected that Optane in App Direct Mode
will be most useful when this inequality is not satisfied \tristan{why?}. As Memory Mode will inevitably
increase the amount of data that can be cached by the storage, it is expected that the overall
performance of all storage devices \tristan{will what?}, even when the inequality is not satisfied.

@@

@@

For our experiments, we used the entire CoRR dataset, made available on DataLad and
applied Participant analysis on it. As in the BigBrain Incrementation, the experiment parallelized 
using GNU Parallel with 25 and 96 processes. The conditions were executed in both App Direct and Memory Mode
using Isilon, local disk, Optane and tmpfs as storage. As participant analysis operates on only a subset of the dataset,
the inequality is satified for all conditions. Each experiment was repeated 3x.
\section{Results}

@@

When observing the total task breakdown (Figure~\ref{fig:stacked-25cpus}), it is
apparent that the differences between the makespans arise from difference in incrementation
time between all devices except Isilon. In addition, while tmpfs and local disk have comparable write times, Optane
has a more significant write time difference. Whereas, Isilon has more significant read and
write time compared to all other devices, it's incrementation time is comparable to that
of local disk. Further inspection of each storage device's Gantt chart (see Figure~\ref{fig:gantt25})
show that increased duration of tasks stagger the execution of each individual tasks. This, in turn,
results in better performance towards the end of the execution, where tasks are significantly more staggered.
Only tmpfs appears to be able to execute each task efficiently with little staggering.
\tristan{In the methods, you should say how time was measured.}
\tristan{Is Isilon in writeback? If so, why is write time so large?}
\subsubsection{96 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{../experiments/bigbrain-incrementation/figures/makespan-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Makespan of \bigbrain Incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:makespan-96cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{../experiments/bigbrain-incrementation/figures/stacked-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{total read/increment/write breakdown of \bigbrain incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:stacked-96cpus}
\end{figure}

@@

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{../experiments/bids-app-example/25cores-sum.pdf}
    \caption{25 processes}\label{fig:bb25}
\end{subfigure}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{../experiments/bids-app-example/96cores-sum.pdf}
    \caption{96 processes}\label{fig:bb96}
\end{subfigure}
\caption{I/O and CPU breakdown of BIDS App Example 25 and 96 processes on all storage devices. 3 repetitions were performed}\label{fig:bbd}
\end{figure*}
\tristan{I/O time might include sleep time if the process is throttled, we should explain how I/O and CPU were measured.}
Unlike the BigBrain incrementation at 25 Processes, the makespan of the BIDS App Example
executing on Optane is longer than local disk (Figure~\ref{fig:bm25}). However, similarly
to the BigBrain Incrementation, tmpfs has a slight lead over local disk and Isilon performs the
worst. The same pattern can be found with 96 process (Figure~\ref{fig:bm96}). When comparing
the executions of 25 and 96 processes, the performance improved by a similar factor on all
storage devices.

@@

While Memory Mode provides a cost-effective solution to increasing the total memory
size, it is slower than App Direct Mode for embarrassingly parallel applications such
as \bigbrain incrementation. This is likely due to the Page Cache being composed of a
combination of both DRAM and Optane in Memory Mode. Since the data produced by the application should be
able to fit within DRAM, writing to a slower bandwidth device such as Optane inevitably
deteriorates performance. As App Direct separates usage of DRAM and Optane, devices only utilize DRAM in this mode.
The only device that performed better in Memory Mode was 
Isilon \tristan{But the difference looks very small}. This may be explained by a larger Page Cache in Memory Mode, enabling Isilon to
store more data in memory prior to initiating synchronous writes to Isilon.

I/O-less task duration also seem to be affected by the execution mode, with those tasks
also performing significantly better in App Direct. A potential explanation for this may be 
that all memory accesses to Optane are slower due to the devices bandwidth, meaning that
tasks that do not perform I/O, but access data loaded into memory are slower in Memory Mode.
This can be seen in Figure~\ref{fig:gantt96} where for local disk, incrementation in App Direct
mode is significantly shorter. Only when the memory bus is overwhelmed does incrementation take
longer for App Direct, however, it is nothing compared to the average duration of the incrementation
tasks in Memory Mode. \tristan{how about throttling?}

>>>>

\begin{abstract}
    Open-access neuroimaging datasets have reached up to $\SI{}{\peta\byte}$ of data, and
    continue to grow. The ability to leverage the entirety of these datasets are
    limited to a restricted number of labs with both the capacity and infrastructure
    to process the data. Whereas BigData engines have significantly reduced
    application performance penalties with respect to data movement, their applied
    strategies are not necessarily practical within neuroimaging workflows where
    intermediary results may need to be materialized to shared storage for post-processing
    analysis. In this paper we evaluate the performance advantage brought by Intel
    Optane DC persistent memory storage for the processing of large neuroimaging 
    datasets using the two available configurations modes: Memory Mode and App Direct Mode.
    We employ a synthetic algorithm on the $\SI{76}{\gibi\byte}$ \bigbrain
    and a real neuroimaging application on the Consortium for Reliability and Reproducibility (CoRR)
    dataset using 25 and 96 parallel
    processes in both cases. Our results show that performance of all available storage
    devices (e.g. tmpfs, local disk and Isilon) is somewhat deteriorated by memory mode
    when compared to App Direct Mode. In contrast, Optane in App Direct Mode was
    found to be slower than local disk, when the load on the bus was high due to sheer
    amount of concurrent processes and short task duration. It is believed that
    Optane would likely be useful in the accessing of typically larger-than-memory
    data in a limited-parallelism manner, such as with visualization applications.
    Further experiments with larger datasets remain to be performed.

@@

Due to storage limitations, only subsets of such neuroimaging repositories 
can be processed in a typical research laboratory. Moreover, as these datasets 
are extremely large and are only increasing in size, they are typically stored 
in higher-capacity, slower storage devices, such as hard disk drives, or external
parallel file systems, such as Lustre. In such conditions, large-scale 
studies in neuroimaging remain limited to labs with access to adequate 
infrastructure.

Furthermore, intermediary data is often required for post-processing analysis. This limits
any performance benefits that can arise from volatile in-memory computing as intermediary
results need to be materialized onto persistent storage. As neuroimaging datasets continue
to increase in size, the movement of data will significantly increase the processing time.

To mitigate the effects of data writes, kernel-based
strategies,such as the writeback cache, have been developed. The writeback cache
allows processes to use the memory as a cache for writing. The cache size,
however,
is configurable, but nevertheless limited. When writes approach the cache's capacity,
processes performing writes start to be throttled. After capacity is reached,
processes can no longer use the cache until all written data on cache is flushed
to the appropriate storage device.

Whereas Operating Systems and popular Big Data engines, such as MapReduce~\cite{mapreduce} and Apache Spark~\cite{spark}, have
incorporated software solutions to limit data transfers (e.g. writeback cache, in-memory computing,
data locality, and lazy evaluation), hardware has also adapted to the growing datasets.
One such improvement is the concept of placing flash storage directly on the Dual
In-line Memory Module (DIMM), thereby reducing the latency of accessing data on 
flash storage devices. While the latency of these devices is improved, the bandwidth
remains the same. However, as noted by \cite{nvdimms}, a severe performance degradation
can be experienced by having memory traffic and I/O placed on the same bus.

Intel Optane DC Persistent Memory storage~\cite{optanebrief} is a recently introduced high-performance
storage technology that also sits on the DIMMs to reduce latency to the device.
Unlike DRAM DIMMs, it can accommodate twice the volume ($\SI{512}{\giga\byte}$) of
storage enabling it to be a more cost effective alternative to DRAM storage.
There are two principle configuration modes, namely Memory Mode and App Direct Mode,
that enable the storage to either be accessed as an extension of available memory 
or as a non-volatile memory storage device.

@@

@@

The server used in our experiments contain $\SI{3}{\tera\byte}$ of Intel Optane Persistent memory, 768GB of DRAM,
Isilon, local disk(SSD).
CPU: Cascade lake
OS: Red Hat 7.1
writeback enabled for local disk (unsure about Isilon)
\valerie{include storage bandwidth table here}

@@

As Memory Mode would affect the processing time regardless of the device
selected, we evaluated Memory Mode on tmpfs, local disk and Isilon. tmpfs
consisted of $\SI{298}{\giga\byte}$ and was located in \texttt{/run/user/61218}.
Local disk comprised of a $\SI{240}{\giga\byte}$ Micron 5100 PRO SSD, of which
$\SI{150}{\giga\byte}$ were mounted on \texttt{/home}. The $\SI{628}{\tera\byte}$
Isilon was mounted as NFS4 to \texttt{/home/users}

Optane was not evaluated separately as it is indistinct from DRAM memory in
Memory Mode
\subsubsection{App Direct Mode}

App Direct Mode enables Optane to be accessed as a high-performance storage device.
Unlike Memory Mode, the OS is able to differentiate between main memory and Optane
storage, treating them as two distinct entities. As Optane is located on
the DIMMs, it exhibits lower latencies than traditional SSDs, while being able
to be accessed like an SSD. Optane does not use the DRAM as cache in App Direct Mode.

The same storage devices were used to evaluate App Direct Mode, with the inclusion
of Intel Optane DC Persistent Memory, of which $\SI{1.5}{\tera\byte}$ of storage
were mounted on \texttt{/pvme-1}.

@@

Using this inequality, we anticipate that whenever the application is within the limits,
the storage selection will make little difference, as storage devices with writeback enabled 
will be able to leverage the memory cache. It is expected that Optane in App Direct Mode
will be most useful when this inequality is not satisfied as storage devices will
not be able to use the writeback cache. As Memory Mode will inevitably
increase the amount of data that can be cached by the storage, it is expected that the overall
performance of all storage devices improve, even when the inequality is not satisfied.

@@

Time is measured within the application using Python's \texttt{time} module 
called before and after any reads, increments or writes.

@@

For our experiments, we used the entire CoRR dataset, made available on DataLad and
applied Participant analysis on it. As in the BigBrain Incrementation, the experiment parallelized 
using GNU Parallel with 25 and 96 processes. The conditions were executed in both App Direct and Memory Mode
using Isilon, local disk, Optane and tmpfs as storage. As participant analysis operates on only a subset of the dataset,
the inequality is satified for all conditions. Each experiment was repeated 3x.

Time in this application was obtained using Linux's \texttt{time(1)} application.
\texttt{Real} was used to measure CPU time, where \texttt{Sys} was used to measure
I/O time.
\section{Results}

@@

When observing the total task breakdown (Figure~\ref{fig:stacked-25cpus}), it is
apparent that the differences between the makespans arise from difference in incrementation
time between all devices except Isilon. In addition, while tmpfs and local disk have comparable write times, Optane
has a more significant write time difference. Whereas, Isilon has more significant read and
write time compared to all other devices, it's incrementation time is comparable to that
of local disk. Further inspection of each storage device's Gantt chart (see Figure~\ref{fig:gantt25})
show that increased duration of tasks stagger the execution of each individual tasks. This, in turn,
results in better performance towards the end of the execution, where tasks are significantly more staggered.
Only tmpfs appears to be able to execute each task efficiently with little staggering.
\subsubsection{96 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{../experiments/bigbrain-incrementation/figures/makespan-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Makespan of \bigbrain Incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:makespan-96cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{../experiments/bigbrain-incrementation/figures/stacked-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{total read/increment/write breakdown of \bigbrain incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:stacked-96cpus}
\end{figure}

@@

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{../experiments/bids-app-example/25cores-sum.pdf}
    \caption{25 processes}\label{fig:bb25}
\end{subfigure}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{../experiments/bids-app-example/96cores-sum.pdf}
    \caption{96 processes}\label{fig:bb96}
\end{subfigure}
\caption{I/O and CPU breakdown of BIDS App Example 25 and 96 processes on all storage devices. 3 repetitions were performed}\label{fig:bbd}
\end{figure*}
Unlike the BigBrain incrementation at 25 Processes, the makespan of the BIDS App Example
executing on Optane is longer than local disk (Figure~\ref{fig:bm25}). However, similarly
to the BigBrain Incrementation, tmpfs has a slight lead over local disk and Isilon performs the
worst. The same pattern can be found with 96 process (Figure~\ref{fig:bm96}). When comparing
the executions of 25 and 96 processes, the performance improved by a similar factor on all
storage devices.

@@

While Memory Mode provides a cost-effective solution to increasing the total memory
size, it is slower than App Direct Mode for embarrassingly parallel applications such
as \bigbrain incrementation. This is likely due to the Page Cache being composed of a
combination of both DRAM and Optane in Memory Mode. Since the data produced by the application should be
able to fit within DRAM, writing to a slower bandwidth device such as Optane inevitably
deteriorates performance. As App Direct separates usage of DRAM and Optane, devices only utilize DRAM in this mode.
The only device that performed better in Memory Mode was 
Isilon, however the difference is minimal.
%This may be explained by a larger Page Cache in Memory Mode, enabling Isilon to
%store more data in memory prior to initiating synchronous writes to Isilon.

I/O-less task duration also seem to be affected by the execution mode, with those tasks
also performing significantly better in App Direct. A potential explanation for this may be 
that all memory accesses to Optane are slower due to the device's bandwidth, meaning that
tasks that do not perform I/O, but access data loaded into memory are slower in Memory Mode. Another explanation might be that the kernel is throttling the processes
as they are producing significant amounts of output, thereby increasing the
incrementation time.
This can be seen in Figure~\ref{fig:gantt96} where for local disk, incrementation in App Direct
mode is significantly shorter. Only when the memory bus is overwhelmed does incrementation take
longer for App Direct, however, it is nothing compared to the average duration of the incrementation
tasks in Memory Mode.
